Development of a Reinforcement Learning Enabled Cattle Tracker Prototype

### 2025-02-14 paper/main.tex
updated paper and a sample power consumption (with the GPS included)

The goal of the agent was to maximize the number of messages sent to the cloud, while minimizing the power consumption of the device.
It is not trivial to determine the reward values for the agent to maximize data collection.
We consider the reward values to be hyperparameters that can be tuned to maximize the data collection and used a Bayesian optimization framework from the the python \verb|scikit-optimize| package to automatically tune the hyperparameters.
The hyper parameters could be nagative and they were:
\begin{itemize}
  \item reward\_power\_loss: the reward for losing power
  \item reward\_power\_multiplier: the reward for having excess power at every time step
  \item reward\_action\_0: the reward for transmitting data
  \item reward\_action\_1: the reward for collecting and storing data
  \item reward\_action\_2: the reward for doing nothing
  \item reward\_message\_count: the reward for the number of messages sent when transmission occurred
\end{itemize}

We ran repeated training sessions over solar gain data collected from the physical prototype that guided us toward the parameter set that maximizes data collection.
This automated approach was less labour intensive than manual tuning and has shown to be effective in optimizing the reward values.

@@

\begin{table}[h!]
  \centering
  \caption{Hardware Inventory}
  \begin{tabular}{|l|l|l|l|}
  \hline
  \textbf{Hardware} & \textbf{Serial Number} & \textbf{Purpose} & \textbf{Specs} \\ \hline
  Arduino MKR WiFi & SN12345 & Connectivity & WiFi-enabled microcontroller board \\ \hline
  Arduino MKR GPS  & SN67890 & Location Tracking & Integrated GPS module \\ \hline
  DFRobot Solar Power Manager & SN11223 & Power Regulation & Manages and distributes solar cell and battery \\ \hline
  Solar Panel 70x100 mm & SN44556 & Energy Generation & 70x100 mm panel, 5W output \\ \hline
  Sparkfun Fuel Gage & SN77889 & Fuel Level Monitoring & Analog sensor for fuel measurement \\ \hline
  LiPo Battery & HXJNLDC & Energy Storage & 3.7V, 2100mAh capacity \\ \hline
  \end{tabular}
  \label{tab:hardware_inventory}
  \end{table}

The prototype-level hardware experiences higher power consumption during sleeping, which results in shorter battery life. 
A side-benefit of this larger power consumption is a faster feedback cycle for development and troubleshooting. 
A custom-build PCB is planned for the production-level hardware to reduce power consumption during sleep and ensure robustness of the final device.

\subsection{Firmware}
The firmware for the device is written in C++ and is responsible for managing the power consumption of the device.
The main loop runs at a 30 minute interval and the device alternates between sleeping and decision making.
At each decision point the device reads the time, the power level of the battery, and the number of messages stored in the device and retrivees a recommended action.
The device then takes the recommended action and updates the power level of the battery.

The device is limited in its memory to store the policy.
To pair down the table, the highest action for each state was selected.
the policy was then stored as values that were bit-shifted to fit into a 32-bit integer.

@@

\section{Implementation}

%TODO: Add sample power consumptions graphs

@@

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.5\textwidth]{./figs/prototype_clean.png}
  \caption{Prototype of the reinforcement learning-enabled cattle tracker. The device consists of a solar panel for power harvesting, a rechargeable battery, and an embedded microcontroller for decision-making and communication.}
  \label{fig:prototype}
\end{figure}

>>>>

The goal of the agent is to maximize the number of messages sent to the cloud, while minimizing the power consumption of the device.
It is not trivial to determine the reward values for the agent to maximize data collection.
We consider the reward values to be hyperparameters that can be tuned to maximize the data collection and used a Bayesian optimization framework from the the python \verb|scikit-optimize| package to automatically tune the hyperparameters.
The hyper parameters could be nagative and they were:
\begin{itemize}
  \item reward\_power\_loss: the reward for losing power
  \item reward\_power\_multiplier: the reward for having excess power at every time step
  \item reward\_action\_0: the reward for transmitting data
  \item reward\_action\_1: the reward for collecting and storing data
  \item reward\_action\_2: the reward for doing nothing
  \item reward\_message\_count: the reward for the number of messages sent when transmission occurred
\end{itemize}

The agent was trained on a simulated environment with the power consumption and generation numbers from the prototype.
At every step, the environment would update the power level of the battery, the time of day, and the number of messages queued on the device depending on the action taken by the agent.
The actual measured battery capacity was 1300 mAh and every 30 minutes the device would consume 5.0 mAh in sleeping.
A power transmission would consume an additional 1.6 mAh and a data collection would consume an additional 0.1 mAh.

Sample solar gain data from 17 days in June in Edmonton was turned into an average sunny day and an average cloudy day. 
The environment would switch between the two days to simulate the weather conditions with an 80\% chance of a cloudy day.
These conditions ensured that there was not sufficient power available to transmit data at every time step and forced the agent to learn strategies to increase message count.

The training run was repeated 4 times each with a different random seed at every iteration of the hyper parameter search.
Once an optimum set of hyperparameters was found, the action-value function was deployed into an h-file that was used in the firmware of the device.
This automated approach was less labour intensive than manual tuning and has shown to be effective in optimizing the reward values.

@@

\begin{table}[h!]
  \centering
  \caption{Hardware List for the Cattle Tracker Prototype}
  \begin{tabular}{|l|l|l|}
  \hline
  \textbf{Hardware} & \textbf{Serial Number} &  \textbf{Description} \\ \hline
  Arduino MKR WiFi 1010 & ABX00023 &  WiFi-enabled microcontroller board \\ \hline
  Arduino MKR GPS Shield & ASX00017 &  Integrated GPS module \\ \hline
  DFRobot Solar Power Manager & TPX00046 &  Manages and distributes solar cell and battery \\ \hline
  Solar Panel & TPX00181 &  75x100x2.9 mm panel, 1W output \\ \hline
  Sparkfun Fuel Gage & 1568-1273-ND & Analog sensor for power measurement \\ \hline
  LiPo Battery & 2528 &  3.7V, 2000mAh capacity \\ \hline
  \end{tabular}
  \label{tab:hardware_inventory}
  \end{table}

Testing on prototype-level hardware showed a high power consumption of 9.9 mA during sleeping due to the presence of multiple LEDs and integrated components that could not be disabled. 
A custom-build PCB with isolation of every component should experience a sleep current on the order of $\mu A$ and is planned for the the next iteration hardware to reduce power consumption during sleep.
An unexpected side-benefit of a prototype with large power consumption is a fast development feedback cycle where the prototype drains the battery in days, rather than weeks.

\subsection{Firmware}
The firmware for the protoype is written in C++ and is responsible for managing the power consumption of the device.
The main loop runs at a 30 minute interval and the device alternates between sleeping and decision making.
At each decision point the device reads the time, the power level of the battery, and the number of messages stored in the device and retrivees a recommended action.
The device then takes the recommended action and updates the power level of the battery.

The device is limited in its memory to store the policy.
To pair down the table, only the highest action for each state was selected.
The policy was then stored as values that were bit-shifted to fit into a 32-bit integer.

@@

@@

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.5\textwidth]{./figs/prototype.png}
  \caption{Prototype of the reinforcement learning-enabled cattle tracker. The device consists of a solar panel for power harvesting, a rechargeable battery, and an embedded microcontroller for decision-making and communication.}
  \label{fig:prototype}
\end{figure}
